Simulated Observations of Young Protoplanetary Discs

### 2013-02-04 paper.tex
added next lot of PC comments

\date{In original form 2013}

@@

\begin{abstract}
The formation and earliest stages of protoplanetary discs are still lacking observational constraints. ALMA will soon revolutionise this field, so it is important to provide predictions and help the interpretation of future high sensitivity and high angular resolution observations. Here we present simulated ALMA observations of a relatively massive (0.39 M$_{\odot}$) self-gravitating disc embedded in a 10\,M$_{\odot}$ envelope with a structure similar to the pre-stellar core L1544. We focus on simple species and conclude that OCS is among the best tracers to measure the disc structure and kinematics. 
\end{abstract}

@@

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. Krasnopolsky et al. 2011; Machida et al. 2011; Braiding \& Wardle 2012). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows, thus observations of the young discs, predicted to have sizes of about 100\,AU and masses as large as 10\% the original core mass (e.g. Joos et al. 2012), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well to filter out the extended emission tracing envelope material.

After the pioneer work of, e.g., Chandler et al. (1995), Brown et al. (2000), Looney et al. (2000), recent interferometric observations have discovered compact embedded discs in a sample of Class 0 (Andr\'e et al. 2000) sources, finding masses between 0.4 and  $>$1 M$_{\odot}$ (Jorgensen et al. 2007, 2009; Enoch et al. 2011). A 130\,AU disc was discovered toward a Class 0 source in Perseus, using Jansky Very Large Array (JVLA) observations of NH$_3$ (Choi et al. 2007). Pineda et al. (2012) observed methyl formate with the Atacama Millimetre and millimetre Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. Persson et al. (2012) observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets (Hartogh et al. 2011). Zapata et al. (2013) used ALMA to observe a disc of size $\sim$50$\,$AU in the other Class 0 source in IRAS 16293-2422 and detected infall towards this disc as inverse P-Cygni profiles in HCN and CH$_3$OH. When ALMA will be completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on theoretical models. As full-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs.

Previous work on simulating observations of gravitationally unstable discs have performed calculations of the continuum radiative transfer in order to look at the structure of the disc and detect disc fragmentation (Ruge et al. 2012) or have used the local thermodynamic equilibrium (LTE) approximation in order to estimate the molecular line emission for an unembedded disc (Krumholtz et al. 2007).

In this paper, we focus our attention to the young self-gravitating discs of Boley \& Durisen (2008), where episodic heating induced by spiral shocks is present, as this may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young ``static'' discs studied by, e.g. Visser et al. (2009, 2011). As shown by Ilee et al. (2011), the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, too high to occur at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of 3D radiative transfer modelling, we perform simulated ALMA observations of the disc studied by Ilee et al. (2011) and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect.\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sect.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

The physical model used to simulate the emission from a young protoplanetary disc is a hybrid model obtained from embedding a gravitationally unstable protoplanetary disc  within an envelope with characteristics similar to the well-studied pre-stellar core L1544, as derived by Keto \& Caselli (2010; hereafter KC2010). The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere (Bonnor 1956; Ebert 1957) with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly dissociated (Keto et al., in preparation), and it is shown in Fig.\,\ref{fig:l1544_model}. \newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/L1544model_used_legend_small.eps}
 \caption{Model of the pre-stellar core L1544 used as the envelope of the young protoplanetary disc in the hybrid model. Showing temperature (red) and dust (green) temperature in kelvin, log number density (blue) in cm$^{-3}$ and inward velocity$\times$100 (cyan) in m$\,$s$^{-1}$. Adapted from Keto \& Caselli (2010) and Keto et al. (2013, in preparation).}
 \label{fig:l1544_model}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT6.eps}
 \caption{{\bf Left:} A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. {\bf Right:} The 3D temperature structure of the disc; regions cooler than 40 degrees are not shown in 3D, demonstrating the narrow central region containing hot material.}
 \label{rhoT} 
\end{figure*}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/columnDensities2.eps}
 \caption{Column densities of OCS, H$_2$CO and HCO$^+$ used in the disc model. Figure adapted from Ilee at al. (2011) [SHOULD WE ALSO ADD C18O COLUMN DENSITY AS WE WANT TO DISCUSS THAT LATER ?? ]}
 \label{Chemistry} 
\end{figure*}

@@

The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the young protoplanetary disc has been plugged in. The disc structure is derived by the hydrodynamic model of Boley(2007) and Boley (2009). The particular model considered here (as well as in Ilee et al. 2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating disc featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The model is sampled over a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$AU in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout both sections model and the opacities were given by the model of dust grains with thick icy mantles and 10$^6\,$yr coagulation from Ossenkopf and Henning (1994). The gas number density and temperature structure are displayed in Figure \ref{rhoT}. \newline

Chemical abundances in the disc were taken from Ilee et. al (2011) which followed the changes of chemical abundances of trace particles moving through the disc as it evolved. The abundances of 125 species related by 1334 reactions were followed through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$AU.  From Ilee et al. (2011) we selected the three species which appear to trace different regions of the disc : the inner 20\,AU (OCS), the inner 40\,AU (H$_2$CO) and the region between $\simeq$40 and 60\,AU (HCO$^+$) and one which traces the entirety of the disc (C$^{17}$O). As explained by Ilee et al. (2011),  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also Caselli et al. 2012). As discussed in the result section (Sec.\,\ref{sec:model_results}), rough guesses have been made based on values measured toward similar objects.

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
%-describe the RT used (LIME) 
%-describe the RT used (LIME) 
The radiative transfer program used, LIME (Brinch \&Hogerheijde 2010), calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in the grid-construction section. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and turbulent velocity are taken from the model. These points are then smoothed by Lloyd's algorithm (Lloyd 1982) in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation \footnote{In three dimensions this means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figures \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in Brinch \&Hogerheijde 2010) to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. \newline
After 20 iterations of updating the level populations the model is ray-traced in order to produce synthetic brightness maps. In order to minimise the artifacts in the output images, resulting from the grid construction, the average of ten separate runs was taken (figure \ref{averages}).

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/grid.eps}%Lime_grid3.eps
 \label{grid}
 \caption{A plot of the points selected by the griding process and the paths down which photons can propagate overlaid on a smoothed density model. The points are more concentrated at small radii and in the densest regions.}
\end{figure}

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime_points_rz_histo2.eps}
 \label{points}
 \caption{A 2D histogram of the point distribution throughout the model. The disc and envelope can be seen as two separate entities which have to be sampled using different point distributions}
\end{figure}

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. The Molecular data used is from the lamda database (Sch\"oier et al. (2005) http://home.strw.leidenuniv.nl/$\sim$moldata/). Lines for the molecules which satisfy these conditions were simulated with LIME for frequencies between 50$\sim$500$\,$GHz, with a focus on lines which could be observed in ALMA band 7 and molecules which are not typically seen in outflows. The results presented in this section represent those lines which show detectable emission/absorption which can be used to trace either spiral structure or rotation.\newline

Transitions focused upon are CO 3-2 (and isotopologues) at 345.8GHz with an upper energy level of 33.2K, HCO$^+$ 3-2 at 267GHZ and 25.7K, OCS 26-25 at 316GHz and 205K and H$_2$CO $4_{04}$-3$_{03}$ at 290GHz and 34.9K {\bf maybe worth including a higher frequency line as well? CS at ~650GHz?}. Other molecules simulated but not shown include HCN, HNC, HNO, HCS$^+$ and CS.\newline

For the purpose of simulating observations the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and is inclined at 30$^\circ$ to edge on. From these observations integrated intensity, intensity weighted velocity and position velocity diagrams through the centre of the model were created.
The simulations were focused upon frequencies within ALMA band 7, selected to give the best trade off between resolution and sensitivity for early ALMA science.
(Note the moment 1 and 0 maps were created by integrating between -12.5 to -0.5 km$\,$s$^{-1}$ and +0.5 to +12.5 km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope, this can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity, moment 1 maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions})\newline

@@

OCS traces only the innermost region of the disc and can be used to examine the central regions (figure \ref{OCS_all}). As OCS is not seen in outflows it can be used to view rotation of the central part of the disc without risk of contamination. As the excitation energy for the high J transitions of OCS is much higher than the other lines under consideration it is only excited in the hottest densest regions. As a result of this the spiral structure is clearer in OCS than in most transitions, with the low density regions between spiral arms showing no emission or absorption(see figure \ref{OCS_all}). The majority of the OCS is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this in the centre of the disc and is an effect of the inclination angle, whereby the line of sight passes through gas above the mid-plane and then through the hole in the centre of the disc.

@@

The abundance of the H$_2$CO molecule is fairly typical of the molecules simulated in Ilee (2011) in that it traces the spiral structure roughly the inner 2/3$^{rd}s$ of the disc. The spiral structure of the disc can be seen clearly in the integrated intensity map and its rotation of detected out to larger radii than is possible with the OCS and is more visible between spiral arms but is otherwise similar in morphology.

@@

Some molecules such as HCO$^+$ trace only the outer regions of the disc (Ilee 2011) and so can be used to look at the extended velocity and physical structure. In these colder less dense regions we see the molecular lines in emission rather than absorption. \newline
%Figure \ref{hcop_pv} shows the rotation curve of the model as shown in figure \ref{velocity} against the HCO$^+$ J=1-0 line emission. It is clear that from observations such as these the rotation curve of a disc

In all the simulations except HCO$^+$ we see molecular lines in absorption throughout the majority of the disc and in some cases a small amount of emission towards the centre. Molecular lines throughout the plane of the disc show up in absorption against the continuum emission of the shock heated, dense mid-plane.

@@

Figure \ref{h2co_chanmap} shows a channel map of the H$_2$CO 4$_{04}$-$3_{03}$ line overlaid on the 1$\,$mm continuum emission. From this it can be seen that the absorption features trace the disc spiral structure whilst line emission emanates from gaps in the disc where there is less continuum to be absorbed.

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/casa_all_30deg_contSub.eps}
 \caption{Integrated intensity maps for the simulated observations of:{\bf Top Left:} H$_2$CO 4$_{04}\rightarrow$3$_{03}$, {\bf Top Right:}C$^{17}$O 3$\rightarrow$2, {\bf Bottom Left:} OCS 26$\rightarrow$25 and {\bf Bottom Right:} HCO$^+$ 3$\rightarrow$2. contours start at 3$\sigma$ and increasing in intervals of 3$\sigma$ for each of the maps except the H$_2$CO which starts at 5$\sigma$ and increases in intervals of 5$\sigma$. Emission is in red contours, absorption in blue.}
\label{mom0_maps}
\end{figure*}

@@

\begin{figure*}
 \includegraphics[width=198mm]{Figures/sim/casa_all_30deg_PV_rotCurve.eps}
 \caption{Position velocity diagrams for each of the species simulated with CASA. The green curve on each diagram is the best fit curve to the azimuthally averaged rotation profile from figure 4.{\bf *****labels of h2co and ocs are wrong way round*****}}
 \label{pvs}
\end{figure*}

@@

 \caption{Spectra for each of the simulated species. In each diagram the three spectra are taken from positions on the Y=0 axis, one is at X=0 (green) and the other 2 are from x=$\pm$ N (plus blue, minus red) where N is different for each species. the values for N are: OCS 10AU, H$_2$CO 20AU, C$^{17}$O 28AU, HCO$^+$ 38AU. {\bf maybe should put stars or something on the mom0 maps to show these locations?}}
 \label{spectra}
\end{figure*}

Figure \ref{spectra} shows spectra from the lines simulated, one from the centre of the image and one on each side of the disc along the Y=0 axis. The most notable thing about these is, despite the fact that the spectra are equally spaced about the centre of the disc, they are not symmetric. Most noticeably in the C$^{17}$O the spectrum from the +X side of the disc is shows significantly more absorption than that of the -X side. In the OCS lines, the two sides are not moving with equal velocities and the line widths and shapes are markedly different. These can be atributed to the non-axissymmetric nature of the model.
%Spectra show stuff, i guess? not really sure if this adds anything to the pv diagrams.

@@

>>>>

\date{Jan 2013}

@@

\begin{abstract}
The formation and earliest stages of protoplanetary discs are still lacking observational constraints. ALMA will soon revolutionise this field, so it is important to provide predictions and help the interpretation of future high sensitivity and high angular resolution observations. Here we present simulated ALMA observations based on radiative transfer modelling of a relatively massive (0.39 M$_{\odot}$) self-gravitating disc embedded in a 10\,M$_{\odot}$ dense core, with structure similar to the pre-stellar core L1544. We focus on simple species and conclude that OCS, C$^{17}$O, H$_2$CO and HCO$^+$ lines together allow measurements of the disc structure and kinematics at all scales. 
\end{abstract}

@@

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. Krasnopolsky et al. 2011; Machida et al. 2011; Braiding \& Wardle 2012; Joos et al. 2013). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,AU and masses as large as 10\% the original core mass (e.g. Joos et al. 2012), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well to filter out the extended emission tracing envelope material.

After the pioneer work of, e.g., Chandler et al. (1995), Brown et al. (2000), Looney et al. (2000), recent interferometric observations have discovered compact embedded discs in a sample of Class 0 (Andr\'e et al. 2000) sources, finding masses between 0.4 and  $>$1 M$_{\odot}$ (J\o rgensen et al. 2007, 2009; Enoch et al. 2011). A 130\,AU disc was discovered toward a Class 0 source in Perseus, using Jansky Very Large Array (JVLA) observations of NH$_3$ (Choi et al. 2007). Pineda et al. (2012) observed methyl formate with the Atacama Millimetre and millimetre Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. Persson et al. (2012) observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets (Hartogh et al. 2011). Zapata et al. (2013) used ALMA to observe a disc of size $\sim$50$\,$AU in the other Class 0 source in IRAS 16293-2422 and detected infall towards this disc as inverse P-Cygni profiles in HCN and CH$_3$OH. When ALMA will be completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on theoretical models. As full-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs.

%Previous work on simulating observations of gravitationally unstable discs have performed calculations of the continuum radiative transfer in order to look at the structure of the disc and detect disc fragmentation (Ruge et al. 2012) or have used the local thermodynamic equilibrium (LTE) approximation in order to estimate the molecular line emission for an unembedded disc (Krumholtz et al. 2007).

Simulated observations of gravitationally unstable discs have already been performed to study the continuum emission and measure the structure and possible fragmentation (Cossins et al. 2010; Ruge et al. 2013). Molecular line emission from massive discs has been simulated by Krumholz et al. (2007), assuming local thermodynamic equilibrium (LTE). In this paper, we focus on the self-gravitating discs of Boley \& Durisen (2008), where episodic heating induced by spiral shocks is present. This may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young ``static'' discs studied by, e.g., Visser et al. (2009, 2011). As shown by Ilee et al. (2011; hereafter I2011), the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, too high to occur at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of 3D radiative transfer modelling, we perform simulated ALMA observations of the disc studied by I2011 and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect.\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sect.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

The physical model used to simulate the emission from a young protoplanetary disc is a hybrid model obtained from embedding a gravitationally unstable protoplanetary disc  within a dense core with characteristics similar to the well-studied pre-stellar core L1544, (Keto \& Caselli 2010; hereafter KC2010). The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere (Bonnor 1956; Ebert 1957) with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated (Caselli et al. 2012; Keto et al. in preparation), and it is shown in Fig.\,\ref{fig:l1544_model}. \newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/L1544model_used_legend_small.eps}
 \caption{Model of the pre-stellar core L1544 used as the envelope of the young protoplanetary disc in the hybrid model. Showing temperature (green) and dust (red) temperature in kelvin, log number density (blue) in cm$^{-3}$ and inward velocity$\times$100 (cyan) in m$\,$s$^{-1}$. Adapted from Keto \& Caselli (2010) and Keto et al. (2013, in preparation).}
 \label{fig:l1544_model}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT6.eps}
 \caption{{\bf Left:} A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. {\bf Right:} The 3D temperature structure of the disc; regions cooler than 40 degrees are not shown in 3D, highlighting the narrow central region containing hot material.}
 \label{rhoT} 
\end{figure*}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/columnDensities2.eps}
 \caption{Column densities of OCS, H$_2$CO and HCO$^+$ used in the disc model. Figure adapted from I2011 [SHOULD WE ALSO ADD C17O COLUMN DENSITY AS WE WANT TO DISCUSS THAT LATER ?? ]}
 \label{Chemistry} 
\end{figure*}

@@

The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the young protoplanetary disc has been plugged in. The disc structure is derived by the hydrodynamic model of Boley (2007) and Boley (2009). The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating disc featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model from which the radiative transefer grid is sampled is a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$AU in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout both sections model and the opacities were adopted from Ossenkops and Henning (1994) and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.\newline

Chemical abundances in the disc were taken from I2011 who followed gas-grain chemical processes during the dynamical evolution of the disc. The abundances of 125 species related by 1334 reactions were calculated through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$AU.  From I2011 we selected the four species which appear to trace different regions of the disc: the inner 20\,AU (OCS), the inner 40\,AU (H$_2$CO),  the region between $\simeq$40 and 60\,AU (HCO$^+$) (see figure \ref{Chemistry}) and one which traces the entirety of the disc (C$^{17}$O). As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also Caselli et al. 2012). As discussed in the result section (Sec.\,\ref{sec:model_results}), rough guesses have been made based on values measured toward similar objects. The molecular data used is from the lamda database (Sch\"oier et al. (2005) http://home.strw.leidenuniv.nl/$\sim$moldata/)

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
%-describe the RT used (LIME) 
%-describe the RT used (LIME) 
The radiative transfer program used is LIME (Brinch \&Hogerheijde 2010),which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in the grid-construction section (section \ref{gridding}). At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm (Lloyd 1982) in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation \footnote{In three dimensions this means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figures \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in Brinch \&Hogerheijde 2010) to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. In order to minimise the artifacts in the output images, resulting from the grid construction, the average of ten separate runs was taken (figure \ref{averages}).

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/grid.eps}%Lime_grid3.eps
 \label{grid}
 \caption{A plot of the points selected by the griding process and the paths down which photons can propagate overlaid on the density distribution (in m$^{-3}$). The points are more concentrated at small radii and in the densest regions.}
\end{figure}

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. We focus on the frequency range available with ALMA, with particular attention to band 7, which offers the best trade off between resolution and sensitivity. The results presented in this section are limited to those lines which show detectable emission/absorption which can be used to trace either spiral structure or rotation.\newline

We considered C$^{17}$O 3-2 at 337.1GHz with an upper energy level of $E_u$=32.3K, HCO$^+$ 3-2 at 267GHZ and $E_u$=25.7K, OCS 26-25 at 316GHz and $E_u$=205K and H$_2$CO $4_{04}$-3$_{03}$ at 290GHz and $E_u$=34.9K. Other molecules simulated but not shown include HCN, HNC, HNO, HCS$^+$ and CS.\newline

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and the disc is inclined at 30$^\circ$ to edge on. From these simulated observations, integrated intensity, intensity weighted velocity and position velocity diagrams through the centre of the model were created.
The simulations were focused upon frequencies within ALMA band 7, selected to give the best trade off between resolution and sensitivity for early ALMA science.
(Note the integrated intensity and intensity weighted velocity maps (figures \ref{OCS_all} - \ref{hcop+_all} and \ref{mom0_maps}) were created by integrating between -12.5 to -0.5 km$\,$s$^{-1}$ and +0.5 to +12.5 km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope, this can be seen in some PV diagrams  (figure \ref{PVs}) as the strong absorption feature at all positions around zero velocity, moment 1 maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions})\newline

@@

OCS traces only the innermost 20$\,$AU of the disc and can be used to examine the central regions (figure \ref{OCS_all}). As OCS is not seen in outflows (e.g. Stanke 07, Van der Tak 03) it can be used to view rotation of the central part of the disc without risk of contamination from emission from shocked material. The OCS line, which has the highest excitation among the selected transitions, well traces the hottest and densest regions of the disc. As a result, the spiral structure is best seen in OCS (26 - 25), as shown in figure \ref{OCS_all}. The majority of the OCS line is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc where the hole in the hot, dense midplane means that there is little continuum to be absorbed.

@@

The abundance of the H$_2$CO molecule is typical of the majority of the molecules simulated in I2011 in that it traces the spiral structure roughly the inner 40$\,$AU of the disc. The spiral structure of the disc can be seen clearly in the integrated intensity map and its rotation of detected out to larger radii than is possible with the OCS (figure \ref{h2co_all}). As with the OCS line, the majority of the H$_2$CO is seen in absorption against the disc mid-plane continuum with the central region seen in emission. Unlike the OCS however, the H$_2$CO extends out to large enough radii to be present in the voids between spiral arms. In these regions the H$_2$CO line can be seen in emission, this can be seen most clearly in the H$_2$CO channel map (figure \ref{h2co_chanmap}).

@@

HCO$^+$ in unique in the molecules simulated in I2011 in that it traces only the outer regions of the disc, and so can be used to look at the extended velocity and physical structure. In figure \ref{hco+_all} we can see that the HCO$^+$ line shows the outer edges of the spiral arms in absorption, but also shows emission from the diffuse gas further out in the disc. This allows the velocity structure of the disc measured out to larger radii.\newline
%Figure \ref{hcop_pv} shows the rotation curve of the model as shown in figure \ref{velocity} against the HCO$^+$ J=1-0 line emission. It is clear that from observations such as these the rotation curve of a disc

%With the exception of HCO$^+$, molecular lines are seen in absorption throughout the majority of the disc and in some cases a small amount of emission towards the centre. Molecular lines throughout the plane of the disc show up in absorption against the continuum emission of the shock heated, dense mid-plane.

@@

Figure \ref{h2co_chanmap} shows a channel map of the H$_2$CO 4$_{04}$-$3_{03}$ line overlaid on the 1$\,$mm continuum emission (figure \ref{continuum}). From this it can be seen that the absorption features trace the disc spiral structure whilst line emission emanates from gaps in the disc where there is less continuum to be absorbed.

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/casa_all_30deg_contSub_nums.eps}
 \caption{Integrated intensity maps for the simulated observations of:{\bf Top Left:} H$_2$CO 4$_{04}\rightarrow$3$_{03}$, {\bf Top Right:}C$^{17}$O 3$\rightarrow$2, {\bf Bottom Left:} OCS 26$\rightarrow$25 and {\bf Bottom Right:} HCO$^+$ 3$\rightarrow$2. contours start at 3$\sigma$ and increasing in intervals of 3$\sigma$ for each of the maps except the H$_2$CO which starts at 5$\sigma$ and increases in intervals of 5$\sigma$. Emission is in red contours, absorption in blue. The numbers refer to the spectra shown in figure \ref{spectra}.}
\label{mom0_maps}
\end{figure*}

@@

\begin{figure*}
 \includegraphics[width=198mm]{Figures/sim/casa_all_30deg_PV_rotCurve.eps}
 \caption{Position velocity diagrams for each of the species simulated with CASA. The green curve on each diagram is the best fit curve to the azimuthally averaged rotation profile from figure 4.}
 \label{pvs}
\end{figure*}

@@

 \caption{Spectra for each of the simulated species. In each diagram the three spectra are taken from positions on the Y=0 axis, one is at X=0 (green) and the other 2 are from x=$\pm$ N (plus blue, minus red) where N is different for each species. the values for N are: OCS 10AU, H$_2$CO 20AU, C$^{17}$O 28AU, HCO$^+$ 38AU. The locations for these spectra are given in figure \ref{mom0_maps}, the red green and blue curves are taken from the locations marked 1, 2 and 3 respectively.}
 \label{spectra}
\end{figure*}

Figure \ref{spectra} shows spectra from the lines simulated, one from the centre of the image and one on each side of the disc along the Y=0 axis. The most notable thing about these is, despite the fact that the spectra are equally spaced about the centre of the disc, they are not symmetric. Most noticeably in the C$^{17}$O the spectrum from the +X side of the disc is shows significantly more absorption than that of the -X side. In the OCS lines, the two sides are not moving with equal velocities and the line widths and shapes are markedly different. These can be atributed to the non-axissymmetric nature of the model.

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime_points_rz_histo2.eps}
 \caption{A 2D histogram of the point distribution throughout the model. The disc and envelope can be seen as two separate entities which have to be sampled using different point distributions}
 \label{points}
\end{figure}
